---
title: "ds202_lab4"
author: "Patrick Origer"
date: "10/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(reshape2)
library(tidyr)
library(dplyr)
library(readxl)
library(ggplot2)

defensive <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2019.xlsx', sheet='Defensive')
offensive <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2019.xlsx', sheet='Biography')
```


1. Cleaning Biography Sheet
```{r}
biography <- biography %>% mutate(Weight=as.numeric(Weight)) %>%
  separate(Height, into = c('Feet', 'Inches'), sep = '-') %>%
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches), Height= Feet * 12 + Inches)

bio2019 <- biography

bioClean <- biography
str(bioClean)
```

1. Cleaning Defensive Sheet
```{r}
defensive <- defensive %>% mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent)) %>%
  mutate(across(where(is.character), as.numeric))

def2019 <- defensive

defClean <- defensive
str(defClean)
```

1. Cleaning Offensive Sheet
```{r}
offensive <- offensive %>% mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent)) %>%
  mutate(across(where(is.character), as.numeric))

off2019 <- offensive

offClean = offensive
str(offClean)
```

2.

```{r}
defPrePivot <- defClean
defClean <- defClean %>% pivot_longer(Tackles_Solo:Pass_PB, names_to = 'stat')
```

```{r}
ggplot(defClean, aes(x = value)) + geom_histogram() + facet_wrap(~stat, scales = 'free_y') +
  labs(title = "Histograms Displaying How Often Each Defensive Skill is Used", x = "Number of Defensive Statistics", y = "Number of Statistics Separated by the Game Played and Player")
```

2. Question 2: The rarest skills seem to be Turnover_FF, Turnover_FR, Turnover_NT, and Tackles_Sack.

```{r}
defQ2 <- defClean
defQ2 <- defQ2 %>% filter(Opponent_Opponent %in% c("Iowa", "Notre Dame"), stat == "Tackles_Solo") %>%
  pivot_wider(names_from = 'Opponent_Opponent', values_from = 'value')

defQ2 <- rename(defQ2, Notre_Dame = 'Notre Dame')

ggplot(defQ2, aes(x = Iowa, y = Notre_Dame)) + 
  geom_point() +
  labs(title = "Solo Tackle Comparison for ISU Defense Against Iowa and Notre Dame", y = "Notre Dame")
```

2. Question 3: ISU had played good defense against Notre Dame.

2. Question 4: Completed below.
```{r}
bioClean <- bioClean %>% separate(Hometown, into = c("Town", "State"), sep = ", ")

head(bioClean)
```

```{r}
bioQ2 <- bioClean %>% group_by(State) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
bioQ2
```

2. Part 5: Above is the number of ISU players from each state.

```{r}
offDatStates <- offClean %>% select(Name)
offDatStates <- distinct(offDatStates)

defDatStates <- defClean %>% select(Name)
defDatStates <- distinct(defDatStates)

defStates <- left_join(defDatStates, bioClean) %>%
  select(Name, State)
offStates <- left_join(offDatStates, bioClean) %>%
  select(Name, State)
```


```{r}
ggplot(defStates, aes(State)) + geom_histogram(stat = "count") + labs(title = "Number of Offensive ISU Players From Each State", y = "Number of Players")
```
```{r}
ggplot(offStates, aes(State)) + geom_histogram(stat = "count") + labs(title = "Number of Defensive ISU Players From Each State", y = "Number of Players")
```

3. Question 1: It seems that there are more ISU offensive players from Iowa than any other state. This is not reflected in defensive players where Florida and Oklahoma players are the most prominent. These discrepancies could be attributed to there being quite a few null values for what states players are from.

```{r}
offClean <- offClean  %>% mutate(Total_YDS = ifelse(is.na(Rushing_YDS), 0, Rushing_YDS) + ifelse(is.na(Receiving_YDS), 0, Receiving_YDS) + ifelse(is.na(Passing_YDS), 0, Passing_YDS)) %>%
  mutate(isPurdy = ifelse(Name == "Purdy, Brock", "Brock Purdy", "Not Brock Purdy"))
```


```{r}
ggplot(offClean, aes(x=Opponent_Opponent, y=Total_YDS)) +
  geom_point(aes(colour = isPurdy)) +
  theme(axis.text.x =element_text(angle=45)) +
  labs(title = "Total Yard Comparison of Brock Purdy and the Rest of the Offense Per Game", x = "Opponent", y = "Total Yards")
```

```{r}
offBP <- offClean %>% group_by(isPurdy, Opponent_Opponent) %>%
  summarise(Mean_Total_YDS = mean(Total_YDS, na.rm = TRUE),
            Sum_Total_YDS = sum(Total_YDS, na.rm = TRUE)) %>%
  arrange(Opponent_Opponent)

offBP

```

```{r}
ggplot(offBP, aes(x=Opponent_Opponent, y=Sum_Total_YDS)) +
  geom_point(aes(colour = isPurdy)) +
  theme(axis.text.x =element_text(angle=45)) +
  labs(title = "Total Yards Gained Per Game for Brock Purdy and the Rest of Offense", x = "Opponent", y = "Summed Total Yards")
```

3. Question 2: It seems that Brock Purdy's performance and the rest of the offense perform consistent with one another. When Brock Purdy earns high amounts of yards in a game, the rest of the offense typically also receives high amounts of yards and vice ce versa.


```{r}
defensive2018 <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2018.xlsx', sheet='Defensive')
offensive2018 <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2018.xlsx', sheet='Offensive')
biography2018 <- read_excel('C:/Users/15154/Downloads/cyclonesFootball2018.xlsx', sheet='Biography')
```

```{r}
biography2018 <- biography2018 %>% mutate(Weight=as.numeric(Weight)) %>%
  separate(Height, into = c('Feet', 'Inches'), sep = '-') %>%
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches), Height= Feet * 12 + Inches)

offensive2018 <- offensive2018 %>% mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent)) %>%
  mutate(across(where(is.character), as.numeric))

offensiveQ3 <- offensive2018 %>% mutate(Total_YDS = ifelse(is.na(Rushing_YDS), 0, Rushing_YDS) + ifelse(is.na(Receiving_YDS), 0, Receiving_YDS) + ifelse(is.na(Passing_YDS), 0, Passing_YDS)) %>%
  group_by(Name) %>%
  summarise(Total_Yards_2018 = sum(Total_YDS))

off2019 <- off2019 %>% mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent)) %>%
  mutate(across(where(is.character), as.numeric))

off2019 <- off2019 %>% mutate(Total_YDS = ifelse(is.na(Rushing_YDS), 0, Rushing_YDS) + ifelse(is.na(Receiving_YDS), 0, Receiving_YDS) + ifelse(is.na(Passing_YDS), 0, Passing_YDS)) %>%
  group_by(Name) %>%
  summarise(Total_Yards_2019 = sum(Total_YDS))

offensiveQ3 <- full_join(offensiveQ3, off2019, by = c("Name" = "Name")) %>%
  drop_na(Total_Yards_2018, Total_Yards_2019) %>%
  mutate(Yardage_Improvement_Ratio = Total_Yards_2019 / Total_Yards_2018, Yards_Difference = Total_Yards_2019 - Total_Yards_2018) %>%
  arrange(desc(Yards_Difference))

offensiveQ3
```


```{r}
defensive2018 <- defensive2018 %>% mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent)) %>%
  mutate(across(where(is.character), as.numeric))

defensive2018 <- defensive2018 %>% mutate(Tackle_Interactions_2018 = Tackles_Solo + Tackles_ASST + Tackles_TFL + Tackles_Sack) %>% group_by(Name) %>%
  summarise(Total_Tackles_2018 = sum(Tackle_Interactions_2018))

def2019 <- def2019 %>% mutate(Tackle_Interactions_2019 = Tackles_Solo + Tackles_ASST + Tackles_TFL + Tackles_Sack) %>%
  group_by(Name) %>%
  summarise(Total_Tackles_2019 = sum(Tackle_Interactions_2019))

defensiveQ3 <- full_join(defensive2018, def2019) %>%
  drop_na(Total_Tackles_2018, Total_Tackles_2019) %>%
  mutate(Tackle_Improvement_Ratio = Total_Tackles_2019 / Total_Tackles_2018, Tackle_Difference = Total_Tackles_2019 - Total_Tackles_2018) %>%
  arrange(desc(Tackle_Difference))

defensiveQ3
```

3. Question 3: For offense, it seems that Charlie Kolar improved his yardage from 2018 to 2019 multiple times over. Brock Purdy possessed the greatest yardage increase of all offense with 1673 more yards in 2019 than 2018.

For Defense, it appears that O'Rien Vance had the greatest improvement in total number of tackle interactions moving from 2018 to 2019. Although, Zach Petersen had the highest tackle improvement ratio. Going from only 4 tackle interactions in 2018 to 48.5 in 2019. 

Each of the players discussed above made large improvements from 2018 to 2019.











